Anomaly detection with kernel density estimation on manifolds

### 2022-04-27 paper/paper.tex
Add example of S1 in appendix

For any \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
Define \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function on \(M\) as
\[
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\]
Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{vol}\) on \(T_pM\), and
\(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
\[
d\exp_p^*vol(u) = \theta_p(u)du.
\]

\hypertarget{example}{%
\paragraph{Example}\label{example}}

Manifold: S2

\(\theta \sim U(0, \pi/2)\)

\(X = \cos(\theta)\)

\$\$

>>>>

For any \(u \in T_pM\), let \(\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})\) and let \(w_i, i=1,\dots,d\) be Jacobi fields along \(\gamma\) such that \(w_i(0)=0\), for all \(i=1,\dots,d\), \(Dw_1/dt(0)=u/\|u\|\) and \(Dw_i/dt(0)\) forms an orthonomal basis of the tangent space \(T_pM\).
Define \(\theta_p(q): T_pM \rightarrow \mathbb{R}\) the volume density function on \(M\) as
\[
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
\]
Rewrite \(\theta_p(u) = \theta_p(\log_p(u))\).
The exponential map \(\exp_p: T_pM \rightarrow M\) induced by pullback of a volume form \(\exp_p^*\textit{Vol}\) on \(T_pM\), and
\(\theta_p\) is its density with respect to the Lebesgue measure of the Euclidean structure on \(T_pM\), which gives in normal coordinates
\[
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
\]

\hypertarget{example-s1}{%
\paragraph{\texorpdfstring{Example: \(S^1\)}{Example: S\^{}1}}\label{example-s1}}

Manifold: \(S^1\) with center 0 and radius \(r=1\)

Polar coordinates: \((r, \theta)\)

Uniformly distributed angle: \(\theta \sim U(0, \pi/2)\) with \(f(\theta) = 2 / \pi\) for \(\theta \in [0, \pi /2]\).

Cartesian coordinates: \((x, y)\) where \(x = \cos(\theta) \in [0, 1]\) and \(y = \sin(\theta) \in [0, 1]\)

Density of \(\theta\): \(f(\theta) = 2 / \pi\)

Density of \(x\): \(f(x) = \frac{2}{\pi \sqrt{1-x^2}}\)

The CDF of X is given by
\[
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
\]
Then the density of X is the derivative of the CDF w.r.t. \(x\)
\[
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
\]
Given \((x, y) \in \mathbb{R}^2\) as sample points from \(S^1\), we could estimate the density of each sample point \(X_i \in S^1, i = 1, \dots, N\) and compare it with its true density \(f(x)\).

We use the density estimator in \textcite{Pelletier2005-vu}.
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{h}),
\]
where \(d=2\), \(h\) is the bandwidth, \(\theta_{X_i}(p)\) is the volume density function at sample point \(X_i\), and \(d_g(p, X_i)\) is the geodesic distance between \(p\) and \(X_i\) along the manifold \(S^1\).

Since the manifold is known (\(S^1\)), we know the geodesic distance is the arc length between two points on the manifold, which is given by \(L(\gamma) = r(\beta - \alpha)\). Therefore, the geodesic distance is
\[
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
\]
As stated in \textcite{Pelletier2005-vu}, in terms of \emph{geodesic normal coordinates} at \(X_i\), the volume density function \(\theta_{X_i}(p)\) equals the determinant of the metric \(g\) expressed in these coordinates at the logarithm map \(\exp_{X_i}^{-1}(p)\), i.e.~\(|g(\exp_p^{-1}(x_i))|\).

{[}When using normal coordinates(orthogonal basis) for tangent vector in \(T_{X_i}S^1\), the exponential map has \emph{geodesic normal coordinates}.{]}

The Riemannian metric \(g\) for \(S^1\) is given by the inner products on the tangent spaces of \(S^1\). For each \(X_i \in M\),
\[
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
\]
is smooth in a local neightborhood \(U\) of \(M\).

Then the volume density function \(\theta_{X_i}(p)\) is the inner product of the tangent vectors produced by the logarithm map from \(X_i\) to \(p\), which is given by
\[
\theta_{X_i}(p) = \frac{1}{\sqrt{1 - X_i^2}}.
\]
Therefore, the estimator for \(p \in S^1\) is given by
\[
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} |h|
^{-d} \frac{1}{\sqrt{1 - X_i^2}} K(|\arccos(p) - \arccos(X_i)|^{T}h^{-1}|\arccos(p) - \arccos(X_i)|).
\]
The bandwidth \(h\) is selected by the plug-in bandwidth selector implemented by the R package \emph{ks}.
